Forecasting Multi-District Elections

### 2019-11-30 outputs/paper/paper.Rmd
Update model

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "Forecasting Multi-District Elections"
thanks: "We thank the Social Research Centre, and in particular Ben Phillips, for graciously providing data. Thank you to Monica Alexander, Lauren Kennedy, John McAndrews, Benjamin Stevens, Michael Donnelly, as well as seminar participants at the University of Toronto Political Behaviour Group and the 2019 POP AusPSA Standing Group Annual Workshop for their assistance and helpful suggestions. Our code and datasets are available at: https://github.com/RohanAlexander/ForecastingMultiDistrictElections. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca. For the latest version please click [here](https://github.com/RohanAlexander/ForecastingMultiDistrictElections/raw/master/outputs/paper/paper.pdf)."
author:
- name: Rohan Alexander
  affiliation: University of Toronto
- name: Patrick Dumont
  affiliation: Australian National University
- name: Patrick Leslie
  affiliation: Australian National University
abstract: "The overall outcome of an election in a multi-district system turns on the number of electoral divisions won by each party. Despite this, traditional political opinion polls tend to focus on broad national-level estimates of support for major parties. In this paper we develop a framework to characterise a useful forecasting model of multi-district elections. We then implement an approach to forecast the number of electoral divisions won by each of the larger political parties in Australia at the 2019 Federal Election. Our approach first uses survey data and multi-level regression with post-stratification (MRP) to estimate electoral-division-level first-preference shares. To account for Australia's use of preferential voting, we then apply a model that we trained on previous election results to estimate electoral-division-level two-party-preferred shares. We find that our approach performs well out-of-sample, and it has additional advantages over traditional polling including improved interpretability, transparency, and better communication of statistical uncertainty. Our framework allows the consistent evaluation of forecasting models of elections in multi-district systems. Finally, our paper also improves our understanding of the circumstances in which MRP is appropriate. Specifically, we find that despite the bias of voter advice applications, and despite the relatively small number of respondents per electoral division for a typical political opinion poll, both can still provide informative results on an electoral-division basis."
keywords: "Multi-district elections, political opinion polls, forecasting, multi-level regression with post-stratification, Australia."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
toc: FALSE
# spacing: double
bibliography: references.bib
biblio-style: "apalike"
endnote: no

@@

@@

In this paper we show that it is possible to use the individual-level responses already collected by political opinion polling firms, along with publicly available datasets and accessible quantitative methods to produce estimates of the number of electoral divisions won by each party. We first establish a framework for evaluating a forecasting model for multi-district elections. It consists of five aspects. These are that a model should: be interpretable and provide reasonable estimates of the number of electoral divisions that each party will win; account for preferences and/or turnout; operate in a reproducible and transparent way; produce results that account for the fact that it is a probability model; and, finally, not be worse than traditional approaches in terms of speed or cost. Within this framework, the method that we implement involves first estimating the first-preference share of each party in each electoral division based on survey data using multi-level regression with post-stratification. We then apply a model that we trained to estimate division-level two-party-preferred shares.
<!-- Seat-specific anomalies can be accounted for by adding additional layers to the model, or by inputting detailed information about that seat. And the model, which is centered around multi-level regression with post-stratification goes some way to ensuring that a handful of odd results do not need to be manually adjusted or removed. -->

@@

## Account for features of the electoral system

@@

[^moreyearsfootnote]: Data on preferences are available since preferential voting was introduced in 1918, for instance via the AustralianElections R package [@AustralianElections] or the eechidna R package [@eechidnacite]. Nonetheless we train the model on four elections' worth of data. This is a balance between over-fitting by using too few elections, and ensuring the relevance of the preferences that we are training on. Additionally, because of the need to convert the results into the electoral divisions used in the 2019 Federal Election  every additional election that we train the preferences model on introduces more error.

@@

Equation \ref{eq:pollingequation} is at an individual-respondent level:
<!-- [^betterinformthedivisionsfootnote][^nestbystatesyounumpyfootnote]  -->
\begin{eqnarray} 
    \mbox{Pr}(\hat{\mbox{FP}}_{i, p=1}) = \mbox{logit}^{-1}\left(\delta_{0} + \delta_1\mbox{gender}_{i} + \delta_2\mbox{age}_{i} + \delta_3\mbox{education}_{i} + \delta_4\mbox{division}_{d[i]}\right)
  \label{eq:pollingequation} 
\end{eqnarray} 
It says that the probability that some person $i$, in electoral division $d$, will vote for the Australian Labor Party depends on their gender, their age-group, their highest level of education, and their division. Each of the demographic variables is estimated as a fixed effect.

The division-effect is modelled as:
\begin{eqnarray} 
    \delta_4\mbox{division}_{d[i]} = \alpha_{d[i]} + \beta_{d[i]}\mbox{previous division result}_{d[i]}
\end{eqnarray} 
That is, we consider each division has a degree of 'inertia', whereby support for the Australian Labor Party in the previous election helps determine what we expect support for the Australian Labor Party to be this election.

@@

<!-- [^betterinformthedivisionsfootnote]: **TODO:** We can better inform the divisions coefficient here, but adding a layer so that it defaults to past vote.  -->

<!-- [^nestbystatesyounumpyfootnote]: **TODO:** Need to nest by state you numpy. You forgot to do this and Lauren noticed. Anyway she was nice about it and didn't make a scene, but also helpfully pointed out that it may be something that would allow us to speak to state-differences in the small state-specific independent that runs e.g. Lambia in Tassie, Xenophon in SA, or ONA/Palmer/etc in QLD. This was also a Michael point. -->

@@

@@

@@

Probabilistic forecasts are notoriously difficult to interpret (for instance, see @morss2010examining, @morss2008communicating, @patt2003using, or @mislavsky2019combining). However, in Australia there is a commonly used framework for describing how 'safe' an electoral division is. One way to better communicate our results is to translate our central estimates into this framework. Table \ref{tab:bothoutcomesbyclass} classifies the two-party-preferred share for both our model estimates and the actual outcomes into commonly used descriptive tranches.

The classifications used by the Australian Electoral Commission are usually used only on actual results, and so we adjust the central classification in a way that is more explicit about uncertainty. Specifically: 'safe loss' means a two-party-preferred share of less than 0.4; 'fairly safe loss' means between 0.4 and 0.44; 'marginal loss' means between 0.44 and 0.47; 'too close to call' means between 0.47 and 0.53; 'marginal win' means between 0.53 and 0.56; 'fairly safe win' means between 0.56 and 0.6; and finally, 'safe win' means a two-party-preferred share greater than 0.6. Table \ref{tab:haveagoyougetago} shows this disaggregation. Using this classification,  estimates based on the Life in Australia dataset tend to be more accurate than those based on the Smartvote Australia dataset.

@@

@@

For reasons that are not clear, much of the political opinion polling literature and practice is focused on trying to forecast results that are only correlated with who wins the election rather than who wins the election. For instance, forecasting the popular vote in the US, rather than the Electoral College, or nation-wide preferences in Australia, Canada, and the UK, rather than seat counts. Our results show that there is no need to do this and that widespread adoption of political opinion polling that focused on seat counts would bring substantial improvements. Nonetheless the reality of the situation is that an approach that was either considerably slower or more costly than the traditional alternatives would not be widely accepted.

@@

One interesting aspect of our approach is the consideration of preferences. Preferential, but compulsory, voting is one way in which the Australian system is different to that of the Canada, the UK and the US. It removes considerable complexity to just consider the first-preferences results. Similarly, many surveys only have information on the state of the respondent, instead of their electoral division. Considering these two variants is worthwhile as they expand the potential use cases of the approach. 
<!-- They are implemented in Appendix \ref{variants}. -->

@@

<!-- # Variants -->

<!-- ## State-based-geography -->
<!-- **TBD** -->

<!-- Redo with similar model, but only take state inputs, not electoral-division. Likely would increase the number of survey inputs available. -->

<!-- ## First-preferences only -->
<!-- **TBD** -->

<!-- Redo just considering first-preferences. Substantial decrease in complexity and uncertainty, but what is the tradeoff in terms of accuracy? -->

>>>>

---
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    fig_caption: true
    number_sections: true
    latex_engine: pdflatex
    template: svm-latex-ms.tex
title: "Forecasting Multi-District Elections"
thanks: "We thank the Social Research Centre, and in particular Ben Phillips, for graciously providing data. Thank you to Monica Alexander and Lauren Kennedy for their assistance. Thank you to John McAndrews, Benjamin Stevens, Michael Donnelly, as well as seminar participants at the University of Toronto Political Behaviour Group and the 2019 POP AusPSA Standing Group Annual Workshop for their helpful comments and suggestions. Our code and datasets are available at: https://github.com/RohanAlexander/ForecastingMultiDistrictElections. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca. For the latest version please click [here](https://github.com/RohanAlexander/ForecastingMultiDistrictElections/raw/master/outputs/paper/paper.pdf)."
author:
- name: Rohan Alexander
  affiliation: University of Toronto
- name: Patrick Dumont
  affiliation: Australian National University
- name: Patrick Leslie
  affiliation: Australian National University
abstract: "The overall outcome of an election in a multi-district system turns on the number of electoral divisions won by each party. Despite this, traditional political opinion polls tend to focus on broad national-level estimates of support for major parties. In this paper we develop a framework to characterise a useful forecasting model of multi-district elections. We then implement an approach to forecast the number of electoral divisions won by each of the larger political parties in Australia at the 2019 Federal Election. Our approach first uses survey data and multi-level regression with post-stratification (MRP) to estimate electoral-division-level first-preference shares. To account for Australia's use of preferential voting, we then apply a model that we trained on previous election results to estimate electoral-division-level two-party-preferred shares. We find that our approach performs well out-of-sample, and it has additional advantages over traditional polling including improved interpretability, transparency, and better communication of statistical uncertainty. Our framework allows the consistent evaluation of forecasting models of elections in multi-district systems. Finally, our paper also improves our understanding of the circumstances in which MRP is appropriate. Specifically, we find that despite the bias of voter advice applications, and despite the relatively small number of respondents per electoral division for a typical political opinion poll, both can still provide informative results on an electoral-division basis."
keywords: "Multi-district elections, political opinion polls, forecasting, multi-level regression with post-stratification, Australia."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
toc: FALSE
# spacing: double
bibliography: references.bib
biblio-style: "apalike"
endnote: no

@@

**Please note that all estimates and models in this paper are *preliminary* and should not be circulated or cited.**

@@

In this paper we show that it is possible to use the individual-level responses already collected by political opinion polling firms, along with publicly available datasets and accessible quantitative methods to produce estimates of the number of electoral divisions won by each party. We first establish a framework for evaluating a forecasting model for multi-district elections. It consists of five aspects. These are that a model should: be interpretable and provide reasonable estimates of the number of electoral divisions that each party will win; account for features of the electoral system; operate in a reproducible and transparent way; produce results that account for the fact that it is a probability model; and, finally, not be worse than traditional approaches in terms of speed or cost. Within this framework, the method that we implement involves first estimating the first-preference share of each party in each electoral division based on survey data using multi-level regression with post-stratification. We then apply a model that we trained to estimate division-level two-party-preferred shares.
<!-- Seat-specific anomalies can be accounted for by adding additional layers to the model, or by inputting detailed information about that seat. And the model, which is centered around multi-level regression with post-stratification goes some way to ensuring that a handful of odd results do not need to be manually adjusted or removed. -->

@@

## Account for electoral system features

@@

[^moreyearsfootnote]: Data on preferences are available since preferential voting was introduced in 1918, for instance using the AustralianElections R package [@AustralianElections] or, for more recent elections, the eechidna R package [@eechidnacite]. Nonetheless we train the model on four elections' worth of data. This is a balance between over-fitting by using too few elections, and ensuring the relevance of the preferences that we are training on. Additionally, because of the need to convert the results into the electoral divisions used in the 2019 Federal Election  every additional election that we train the preferences model on introduces more error.

@@

Equation \ref{eq:pollingequation} is at an individual-respondent level:
<!-- [^aessfootnote][^nestbystatesyounumpyfootnote][^alexfootnote] -->
\begin{eqnarray} 
    \mbox{Pr}(\hat{\mbox{FP}}_{i, p=1}) = \mbox{logit}^{-1}\left(\delta_{0} + \delta_1\mbox{gender}_{i} + \delta_2\mbox{age}_{i} + \delta_3\mbox{education}_{i} + \delta_4\mbox{division}_{d[i]}\right)
  \label{eq:pollingequation} 
\end{eqnarray} 
It says that the probability that some person $i$, in electoral division $d$, will vote for the Australian Labor Party depends on their gender, their age-group, their highest level of education, and their division. Each of the demographic variables is estimated as a fixed effect.

The division-effect is modelled as:
\begin{eqnarray} 
    \delta_4\mbox{division}_{d[i]} = \alpha_{d[i]} + \beta_{d[i]}\mbox{previous division result}_{d[i]}
\end{eqnarray} 
That is, we consider each division has a degree of 'inertia', whereby support for the Australian Labor Party in the previous election helps determine what we expect support for the Australian Labor Party to be this election. One alternative here would be to consider an individual's past vote. However, in the absence of linked or panel data voter's recall of past vote has been found to be inaccurate, for instance @vanElsas2013 finds inaccuracy due to forgetfulness, nonattitudes, and cognitive bias.

@@

<!-- [^aessfootnote]: **TODO:** Can we inform gender and age and education based on AES 2016 results? -->

<!-- [^nestbystatesyounumpyfootnote]: **TODO:** Need to nest by state. This would also allow us to speak to state-differences in the small state-specific independent that runs e.g. Lambia in Tassie, Xenophon in SA, or ONA/Palmer/etc in QLD. -->

<!-- [^alexfootnote]: **TODO:** Alex wants justification for the coefficient selections. Maybe could look at AES or something. -->

@@

Here we discuss our core results. Variations and alternatives are discussed in  Appendix \ref{variants}. This includes, only generating state estimates, examining first-preferences, and considering alternative preferences models.

@@

The Life in Australia first-preferences estimates are fairly similar across the divisions. This may be due to the large amount of 'sharing' that is happening in the model, due to the relatively small sample size on an electoral-division basis. The Smartvote Australia first-preferences estimates are more appropriately distributed. However there is a clear over-estimate of the Australian Labor Party share.

@@

Probabilistic forecasts are notoriously difficult to interpret (for instance, see @morss2010examining, @morss2008communicating, @patt2003using, or @mislavsky2019combining). However, in Australia there is a commonly used classification system for describing how 'safe' an electoral division is. One way to better communicate our results is to translate our central estimates into these classifications. Table \ref{tab:bothoutcomesbyclass} groups the two-party-preferred share for both our model estimates and the actual outcomes into descriptive classifications based on this classification system.

The classifications used by the Australian Electoral Commission are usually used only on actual results, and so we adjust them in a way that better considers uncertainty. Specifically: 'safe loss' means a two-party-preferred share of less than 0.4; 'fairly safe loss' means between 0.4 and 0.44; 'marginal loss' means between 0.44 and 0.47; 'too close to call' means between 0.47 and 0.53; 'marginal win' means between 0.53 and 0.56; 'fairly safe win' means between 0.56 and 0.6; and finally, 'safe win' means a two-party-preferred share greater than 0.6. Table \ref{tab:haveagoyougetago} shows this disaggregation. Using this classification,  estimates based on the Life in Australia dataset tend to be more accurate than those based on the Smartvote Australia dataset.

@@

In this section we evaluate our approach fits within the context of the framework that we established. We then compare the survey inputs and their appropriateness for the task that we have attempted. Finally, we go into the weaknesses and limitations of our approach, and conclude with some next steps.

@@

For reasons that are not clear, much of the political opinion polling literature and practice is focused on trying to forecast results that are only correlated with who wins the election rather than who wins the election. For instance, forecasting the popular vote in the US, rather than the Electoral College, or nation-wide preferences in Australia, Canada, and the UK, rather than seat counts. @McNair1983 discusses the history of polling in Australia in terms of what was done, rather than why. Our results show that there is no need to do this and that widespread adoption of political opinion polling that focused on seat counts would bring substantial improvements. Nonetheless the reality of the situation is that an approach that was either considerably slower or more costly than the traditional alternatives would not be widely accepted.

@@

One interesting aspect of our approach is the consideration of preferences. Preferential, but compulsory, voting is one way in which the Australian system is different to that of the Canada, the UK and the US. It removes considerable complexity to just consider the first-preferences results. Similarly, many surveys only have information on the state of the respondent, instead of their electoral division. Considering these two variants is worthwhile as they expand the potential use cases of the approach and they are implemented in Appendix \ref{variants}.

@@

# Variants

## State-based-geography
**TBD**

Redo with similar model, but only take state inputs, not electoral-division. Likely would increase the number of survey inputs available.

## First-preferences only
**TBD**

Redo just considering first-preferences. Substantial decrease in complexity and uncertainty, but what is the tradeoff in terms of accuracy?

## Alternative preferences
**TBD**

Patrick wants to see what happens if we use the actual 2019 preferences.
